Efficient forwards-time simulation with ancestral recombination graphs

### 2017-10-25 forwards_paper.tex

To record the genealogical history of a simulation
we need to record two things for each new chromosome:
the birth time,
and the endpoints and parental IDs of each distinctly inherited segment.
This is recorded easily, without further processing,
in the tables described above.

For concreteness, here we write out in pseudocode
how to run a neutral Wright--Fisher simulation
with overlapping generations
that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L,
on which for simplicity we assume there is exactly one crossover per generation.
The probability of death per individual each generation is \texttt{death\_prob},
and the whole-chromosome mutation rate per individual is \texttt{mut\_rate}.

\paragraph{Initialize:}
We will build the tables \nodes, \edgesets, \sites, and \mutations,
and keep track of the IDs of the current generation in the list \texttt{pop}.
\begin{lstlisting}[language=Python]
for i in 0:N-1:
    nodes.add_row(time=T)
    pop[i] = i
\end{lstlisting}
(Alternatively, the tables could be initialized by the results of a coalescent simulation.)

\paragraph{Iterate:}
We then step through the generations,
using the functions \texttt{random\_mutation} and \texttt{random\_allele}
to choose positions and alleles for mutations, respectively:
\begin{lstlisting}[language=Python]
for t in 1:T:
    for i in 0:N-1:
        if random.uniform() > death_prob:
            new_pop[i] = pop[i]
        else:
            u = nodes.num_rows  # the ID of the new individual
            new_pop[i] = u
            nodes.add_row(time = T-t)
            a = random.sample(pop)
            b = random.sample(pop)
            bp = random.sample(0:L)
            if bp > 0:
                edgesets.add_row(left=0, right=bp,
                                  parent=a, children=(u,))
            if bp < L:
                edgesets.add_row(left=bp, right=L,
                                  parent=b, children=(u,))
            num_muts = random.poisson(mut_rate)
            for j in 1:num_muts:
                pos = random_mutation()
                if pos not in sites.position:
                    s = sites.num_rows
                    sites.add_row(position=pos,
                           ancestral_state=random_allele())
                else:
                    s = sites.position.index(pos)
                mutations.add_row(site=s, node=u,
                           derived_state=random_allele())
    pop = new_pop
\end{lstlisting}

\paragraph{Finalize:}
to obtain a tree sequence, we need only transform the tables to the required format:
\begin{lstlisting}
sort_tables(nodes, edgesets, sites, mutations)
simplify_tables(nodes, edgesets, sites, mutations)
\end{lstlisting}
which can then be loaded into a tree sequence with \msprime.
(These two functions operate on the tables in place.)

>>>>

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
with overlapping generations that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we assume there is exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.

Let $\randomuniform(A)$ to be an element of the set $A$ chosen
uniformly at random (note that each instance of $\randomuniform(A)$ within
an algorithm listing represents an independent draw from the set $A$).

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Runs a forwards time Wright--Fisher simulation of a populations with $N$
individuals with a probability of death per generation $\delta$ and a chromosome of length
$L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
in the vector $\tau$, and $E$ will contain a set of $(\ell, r, p, c)$ tuples
describing the recorded edges.
}

\algstep{W1.}{Initialisation.}{
 For $0 \leq j < N$, set $P_j \leftarrow j$ and $\tau_j \leftarrow T$.
Set $E\leftarrow \emptyset$ and $n \leftarrow N$.
}

\algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
Set $t \leftarrow t - 1$ and $j \leftarrow 0$.
}

\algstep{W3.}{Individual loop.}{ If $j = N$ set $P \leftarrow P'$ and go to \algref{W2}.
}

\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) > \delta$, set $P'_j \leftarrow P_j$
    and go to \algref{W7}.
}

\algstep{W5.}{New node.}{Set $P'_j \leftarrow n$ and $\tau_n \leftarrow t$.
}

\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

\algstep{W7.}{Record edges}{Set $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x, L, P_b, n) \right\}$
    and $n \leftarrow n + 1$.
}

\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W4}.
}

\jk{I haven't checked this carefully. If we used 1-based indexes it might make things a bit
simpler}.
\end{taocpalg}

Algorithm~\algref{W} is a very simple implementation of a Wright--Fisher model
intended to illustrate how we can record the nodes and edges of a tree sequence
forwards in time. We begin in~\algref{W1} by allocating our initial population $P$
and creating $N$ nodes with birth time $T$ generations ago, recorded in the
vector $\tau$. The set $E$ is used to store the edges that we output during the
simulation, and $n$ is the number of nodes created so far (and so, the ID of
the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
an individual $P_j$ has died in this generation. If it has, we replace it in
steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
individual. When an individual in the population dies, we first allocate
a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
parents $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $n$
from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
$n$, ready to represent the next new node.

\begin{enumerate}
\item Explain how to add mutational information. We can also do this afterwards on the
output trees if they are nuetral.
\item Explain that we can input these trees into an mpsrime tree sequence and they will
work fine. Perhaps a diagram to show the kind of trees we get? (Multiroots, lots of unary nodes.
Can use the msprime SVG API to draw these?) We could refer ahead to figure 2
instead though.
\item However, there is huge amount of redundancy in these. The amount of information stored
grows linearly with time.

\end{enumerate}
